Enforced Invariant Confluence

### 2018-03-26 doc/notes_v1/alternativesproofs.tex
Renaming various docs.

>>>>

\begin{claim}\label{clm:iconfluence-not-implies-1-iconvergence}
  \iconfluence{} $\centernot\implies$ 1-\iconvergence{}.
\end{claim}
\begin{proof}
  Let
    $I \defeq 0 \leq x \leq 2$ and
    $c = \impif{read(x)=0}{add(x,1)}{}$.
  First, we show that $\set{c}$ is \iconfluent{} with respect to $I$. Consider
  an arbitrary
    $D$,
    $B_1$, and
    $B_2$
  where
    $I(D)$,
    $I(B_1@D)$, and
    $I(B_2@D)$.
  We perform a case analysis on $D$.
  \begin{itemize}
    \item \textbf{Case 1: $D = 1 \lor D = 2$.}
      In this case, $D \circ \denote{B_1}(D) \circ \denote{B_2}(D)$ is
      guaranteed to equal $D$ because $c$ does not affect $D$ at all. By
      assumption, $D$ satisfies the invariant.
    \item \textbf{Case 1: $D = 0$.}
      If $D = 0$, then $D \circ \denote{B_1}(D)$ is either $0$ (if $B_1$ is
      empty) or $1$ if $B_1$ is not empty. A symmetric argument holds for
      $B_2$. Thus, $D \circ \denote{B_1}(D) \circ \denote{B_2}(D)$ is either 0,
      1, or 2, all of which satisfy $I$.
  \end{itemize}

  Next, we show that $\set{c}$ is not 1-\iconvergent{} with respect to $I$. Let
  $D = 1$ and $D' = D'' = 0$.
    $I(D)$,
    $I(D \circ \denote{c}(D'))$,
    $I(D \circ \denote{c}(D''))$,
  but
    $D \circ \denote{c}(D') \circ \denote{c}(D'') = 3$
  does not satisfy the invariant.
\end{proof}

### 2018-03-26 doc/notes_v1/formalism.tex
Renaming various docs.

A \emph{database state} is a total function $D: \var \to \ints$.  We say a
database state $D$ \emph{satisfies invariant} $I$, denoted $I(D)$, if $I$
evaluates to true after all variables in $I$ have been replaced by their
mapping in $D$. Let $\dbs$ denote the set of all database states.

A \emph{transaction} is a partial function $t: \var \rightharpoonup \ints$.
\emph{Applying a transaction} $t$ to database state $D$, denoted by $D \circ
t$, produces a new database state defined as follows:
\[
  (D \circ t)(x) \defeq \begin{cases*}
    D(x) + t(x) & $x \in \dom{}(t)$ \\
    D(x)        & otherwise
  \end{cases*}
\]
Note that transaction application is commutative. That is for all database
states $D$ and transactions $t_1$ and $t_2$, $D \circ t_1 \circ t_2 = D \circ
t_2 \circ t_1$.

A transaction chain $C$ created from a set of transactions $T$ is a sequence of
transactions $t_1, \ldots, t_n$ where $t_i \in T$ for $1 \leq i \leq n$. Let $D
\circ C$ be syntactic sugar for $D \circ t_1 \circ \cdots \circ t_n$. We say
$C$ \emph{satisfies invariant} $I$ starting at $D$, denoted $I(C @ D)$, if for
every prefix $C'$ of $C$, $D \circ C'$ satisfies invariant $I$.

A set of transactions $T$ is \iconfluent{} with respect to invariant $I$ if for
all database states $D$ and chains $C_1$ and $C_2$ created from $T$, if $D$
satisfies $I$, $C_1$ satisfies $I$ starting at $D$, and $C_2$ satisfies $I$
starting at $D$, then $D \circ C_1 \circ C_2$ satisfies $I$.
\todo{
  This isn't quite the definition of \iconfluence{} Bailis uses. Ponder whether
  this simpler definition is okay.
}

>>>>

A \emph{database state} is a total function $D: \var \to \ints$.  We say a
database state $D$ \emph{satisfies invariant} $I$, denoted $I(D)$, if $I$
evaluates to true after all variables in $I$ have been replaced by their
mapping in $D$.
%
For example, if $\var = \set{x, y}$, then $D_1 = \set{x:1, y:2}$ and $D_2 =
\set{x:42, y:99}$ are both databases. If $I \defeq x \geq 10$, then $D_2$
satisfies $I$ (i.e. $I(D_2)$) but $D_1$ does not (i.e. $\lnot I(D_1)$).

A \emph{transaction} is a partial function $t: \var \rightharpoonup \ints$.
\emph{Applying a transaction} $t$ to database state $D$, denoted by $D \circ
t$, produces a new database state defined as follows:
\[
  (D \circ t)(x) \defeq \begin{cases*}
    D(x) + t(x) & $x \in \dom{}(t)$ \\
    D(x)        & otherwise
  \end{cases*}
\]
For example, $t_1 = \set{x: 42}$ is the transaction which increments $x$ by 42.
$t_2 = \set{x: 1, y:-2}$ is the transaction which increments $x$ by 1 and
decrements $y$ by $-2$. If $D = \set{x: 100, y:100}$, then $D \circ t_1 =
\set{x:142, y:100}$ and $D \circ t_2 = \set{x:101, y:98}$.

Note that transaction application is commutative. That is for all database
states $D$ and transactions $t_1$ and $t_2$, $D \circ t_1 \circ t_2 = D \circ
t_2 \circ t_1$. Also note that transactions are \emph{blind} in that they do
not read from the database. We'll relax this assumption in \secref{imp}, but
for now we're keeping things simple.

A transaction chain $C$ created from a set of transactions $T$ is a sequence of
transactions $t_1, \ldots, t_n$ where $t_i \in T$ for $1 \leq i \leq n$. Let $D
\circ C$ be syntactic sugar for $D \circ t_1 \circ \cdots \circ t_n$. We say
$C$ \emph{satisfies invariant} $I$ starting at $D$, denoted $I(C @ D)$, if for
every prefix $C'$ of $C$, $D \circ C'$ satisfies invariant $I$. Note that $I(C
@ D)$ implies $I(D)$ (the empty prefix of $C$) and $I(D \circ C)$ (the entire
prefix of $C$).

A set of transactions $T$ is \emph{\iconfluent{}}\footnote{An astute reader
will notice that this isn't actually the definition of \iconfluence{} presented
in \cite{bailis2014coordination}. It's a simpler one. Rest assured, this
simpler definition will be hard enough to deal with. We'll explore the
ramifications of this simplification in \secref{alie}.} with respect to
invariant $I$ if for all database states $D$ and chains $C_1$ and $C_2$ created
from $T$, if $D$ satisfies $I$, $C_1$ satisfies $I$ starting at $D$, and $C_2$
satisfies $I$ starting at $D$, then $D \circ C_1 \circ C_2$ satisfies $I$.

For example, let $\var = \set{x}$ and consider a transactions $t_+ = \set{x:
1}$ and $t_- = \set{x: -1}$ which increment and decrement $x$ respectively. Let
$I \defeq x > 0$ be the invariant that the value $x$ in the database is
positive.
%
$\set{t_+}$ is \iconfluent{} with respect to $I$ because given any database $D$
and chains $C_1$ and $C_2$, if $x$ is positive in $D$, then $x$ will be
certainly be positive in $D \circ C_1 \circ C_2$.

On the other hand, $\set{t_-}$ is not \iconfluent{} with respect to $I$. We can
prove this with a simple counterexample. Let
  $D = \set{x:2}$,
  $C_1 = t_-$, and
  $C_2 = t_-$.
Note that $D$ and $D \circ t_1 = \set{x:1}$ satisfy the invariant, so $I(D)$,
$I(C_1 @ D)$, and $I(C_2 @ D)$. However, $D \circ C_1 \circ C_2 = \set{x: 0}$
does not satisfy the invariant. Note that this is just a formalization of the
proof we saw in \figref{decrement}.

### 2018-03-26 doc/notes_v1/solution.tex
Renaming various docs.

\section{A Solution}\label{sec:solution}
This section posed the question: how do we write an algorithm to decide whether
a set of PN-Counter increment/decrement transactions are \iconfluent{} with
respect to an invariant expressed as a formula of linear equations and
inequalities?  Now that we're armed with an intuitive
(\secref{counter-overview}) and formal (\secref{counter-question})
understanding of the problem, a geometric interpretation
(\secref{counter-geometry}), and a useful lemma (\secref{counter-one}), we're
in a perfect position to present a solution!

@@

  \item
    \eqref{eq:i1} is $I$ with every variable $x$ replaced with $x + dx_1$.
    Intuitively, it predicates our implication on the assumption that if we
    apply the first transaction, our database still satisfies the invariant.
    \eqref{eq:i2} is symmetric to \eqref{eq:i1} but for the second transaction
    instead of the first.

  \item
    \eqref{eq:i12} is $I$ with every variable $x$ replaced with $x + dx_1 +
    dx_2$. Intuitively, it says that if we apply both the transactions, we
    continue to satisfy our invariant.
\end{itemize}

>>>>

\section{A Solution}\label{sec:solution}
This paper posed the question: how do we write an algorithm to decide whether a
set of PN-Counter increment/decrement transactions are \iconfluent{} with
respect to an invariant expressed as a formula of linear equations and
inequalities?  Now that we're armed with an intuitive (\secref{counters}) and
formal (\secref{formalism}) understanding of the problem, a geometric
interpretation (\secref{geometry}), and a useful lemma (\secref{oneisenough}),
we're in a perfect position to present a solution!

@@

  \item
    \eqref{eq:i1} is $I$ with every variable $\alpha$ replaced with $\alpha +
    d\alpha_1$.  Intuitively, it predicates our implication on the assumption
    that if we apply the first transaction, our database still satisfies the
    invariant.  \eqref{eq:i2} is symmetric to \eqref{eq:i1} but for the second
    transaction instead of the first.

  \item
    \eqref{eq:i12} is $I$ with every variable $\alpha$ replaced with $\alpha +
    d\alpha_1 + d\alpha_2$. Intuitively, it says that if we apply both the
    transactions, we continue to satisfy our invariant.
\end{itemize}

### 2018-03-26 doc/notes_v1/vision.tex
Renaming various docs.

\section{Research Vision}

At a very high level, this research tackles the same problems tackled by many
existing research projects like BOOM Analytics \cite{alvaro2010boom}, Bloom
\cite{alvaro2011consistency}, Bloom$^L$ \cite{conway2012logic}, TARDiS
\cite{crookstardis}, and IPA \cite{holt2016disciplined}:

\begin{quotation}
  \emph{How can we make writing weakly consistent programs more principled?}
\end{quotation}

BOOM Analytics, Bloom, and Bloom$^L$ use data-centric declarative programming
and program analysis based on the CALM theorem. TARDiS exposes the entire
history of divergent computations to users in a git-like system to empower them
to programmatically resolve conflicts. IPA uses a type system to ensure weakly
consistent data doesn't leak into strongly consistent data.

\begin{todoitemize}
  \item \todo{Cite and summarize \cite{balegas2015putting},
    \cite{roy2015homeostasis}, \cite{li2012making}, \cite{li2014automating},
    and \cite{gotsman2016cause}.}
  \item \todo{Consider creating a related works section.}
\end{todoitemize}

This system builds off existing research on CRDTs and \iconfluence{} to provide
a library of CRDTs, a language to express invariants over those CRDTs, the
ability to transactionally update the replicated CRDTs, and a decision
procedure to algorithmically determine whether the transactions are
\iconfluent{} with respect to the invariants, ensuring coordination free
execution if they are and potentially giving illustrative counterexamples when
they are not. A simple example of psuedocode written with this system may look
something like this:

>>>>

\section{Research Vision}
This project builds off existing research on CRDTs and \iconfluence{} to
provide a library of CRDTs, a language to express invariants over those CRDTs,
the ability to transactionally update the replicated CRDTs, and a decision
procedure to algorithmically determine whether the transactions are
\iconfluent{} with respect to the invariants, ensuring coordination free
execution if they are and potentially giving illustrative counterexamples when
they are not. A simple example of psuedocode written with this system may look
something like this:

### 2018-03-26 doc/notes_v2/sufficient_conditions.tex
Renaming various docs.

Here are some more examples of transactions $T$ and invariants $I$ that are
\sTIconfluent{} but not \Iclosed{}.

\todo{Think of more compelling examples.}

@@

>>>>

Here are some more examples. \todo{Think of more compelling examples.}

@@

\begin{example}
  Here are some of the example objects and invariants
  in~\cite{bailis2014coordination}. We begin with invariants over G-Sets $A$
  and $B$. Fortunately, a good number of invariants are \Iclosed{}.

  \begin{center}
    \begin{tabular}{|l|l|p{3in}|}
      \hline
      $I$ & \Iclosed{} & Description \\\hline\hline
      %
      $\forall x \in A.\ \phi(x)$ &
      yes &
        Invariants of this form include things like attribute equality and
        attribute inequality.
        \\\hline
      $\exists x \in A.\ \phi(x)$ &
      yes &
        Invariants of this form include things like non-emptiness constraints.
        \\\hline
      $\forall x \in A.\ \forall y \in A.\ \phi(x, y)$ &
      no &
        Invariants of this form include things like uniqueness constraints
        (e.g. $\forall x \in A.\ \forall y \in A.\ x \neq y \implies key(x)
        \neq key(y)$).
        \\\hline
      $\forall x \in A.\ \forall y \in B.\ \phi(x, y)$ &
      no &
        Invariants of this form include things like disjointness (e.g. $\forall
        x \in A. \forall y \in B.\ x \neq y$).
        \\\hline
      $\forall x \in A.\ \exists y \in A.\ \phi(x, y)$ &
      yes &
        Invariants of this form include things like foreign key constraints
        into a single relation. For example, if we have an \texttt{Employee(id,
        bossid)} relation, we might want the invariant $\forall (i_1, b_1) \in
        E.\ \exists (i_2, b_2) \in E.\ b_1 \neq \bot \implies b_1 = i_2$.
        \\\hline
      $\forall x \in A.\ \exists y \in B.\ \phi(x, y)$ &
      yes &
        Invariants of this form include things like foreign key constraints.
        \\\hline
      $\exists x \in A.\ \forall y \in A.\ \phi(x, y)$ &
      no &
        \todo{Figure out useful interpretation of these formula.}
        \\\hline
      $\exists x \in A.\ \forall y \in B.\ \phi(x, y)$ &
      no &
        \todo{Figure out useful interpretation of these formula.}
        \\\hline
      $\exists x \in A.\ \exists y \in A.\ \phi(x, y)$ &
      yes &
        \todo{Figure out useful interpretation of these formula.}
        \\\hline
      $\exists x \in A.\ \exists y \in B.\ \phi(x, y)$ &
      yes &
        \todo{Figure out useful interpretation of these formula.}
        \\\hline
    \end{tabular}
  \end{center}

  For 2P-Sets $A$ and $B$, we see that things get a bit more grim.
  \begin{center}
    \begin{tabular}{|l|l|}
      \hline
      $I$ & \Iclosed{} \\\hline\hline
      %
      $\forall x \in A.\ \phi(x)$ & yes \\\hline
      $\exists x \in A.\ \phi(x)$ & no \\\hline
      $\forall x \in A.\ \forall y \in A.\ \phi(x, y)$ & no \\\hline
      $\forall x \in A.\ \forall y \in B.\ \phi(x, y)$ & no \\\hline
      $\forall x \in A.\ \exists y \in A.\ \phi(x, y)$ & no \\\hline
      $\forall x \in A.\ \exists y \in B.\ \phi(x, y)$ & no \\\hline
      $\exists x \in A.\ \forall y \in A.\ \phi(x, y)$ & no \\\hline
      $\exists x \in A.\ \forall y \in B.\ \phi(x, y)$ & no \\\hline
      $\exists x \in A.\ \exists y \in A.\ \phi(x, y)$ & no \\\hline
      $\exists x \in A.\ \exists y \in B.\ \phi(x, y)$ & no \\\hline
    \end{tabular}
  \end{center}
  \note{%
    This would be a good table to demonstrate that \Iclosure{} is not only
    unnecessary, but also uncommon. In practice, not a lot of things are going
    to be \Iclosed{}. This would motivate our push beyond these simple checks.
    But then again, not a lot of things are going to be invariant-confluent
    either, so it's not a super strong argument.
  }
\end{example}
